---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# demspaces.dev

<!-- badges: start -->
[![R-CMD-check](https://github.com/vdeminstitute/demspaces/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/vdeminstitute/demspaces/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/vdeminstitute/demspaces/branch/main/graph/badge.svg)](https://app.codecov.io/gh/vdeminstitute/demspaces?branch=main)
<!-- badges: end -->

**demspaces** contains helper functions for the Democratic Spaces Barometer project. 

## Installation

Since this package is specific to the democratic spaces project, it will not end up on CRAN. Install from [GitHub](https://github.com/) instead, with:

``` r
# install.packages("remotes")
remotes::install_github("vdeminstitute/demspaces")
```

## Example

```{r}
library("demspaces")

data("states")

mdl   <- ds_logistic_reg("v2x_veracc_osp", states)
preds <- predict(mdl, new_data = states)
head(preds)
```

## Meta

There is a Makefile at the root level of this repo to help with stuff. E.g. to open that static docs:

```bash
make docs
make opendocs
```

To add new models, see the `demspaces/add_new_model.R` script. It will use a template to setup the R and test files. 

