---
title: "Create DV data"
output: 
  github_document:
    toc: true
---

This script applies the cutpoints to the raw DV indicators to create the actual targets that we will use for the models, i.e. indicators for an up or down movement over the 2 years following the year in a row/observation. These targets are denoted with a "_next2" suffix, as well as the direction ("up", "down") in which they go.

Inputs:

- `output/dv_data_1968_on.csv`
- `output/cutpoints.csv`

Outputs:

The main output is:

- `output/dv-data.rds`

The script also writes CSV files for the target outcome variables, so that changes are easier to identify on git. 

- `output/dv-lists/[vdem indicator].csv`



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readr)
library(states)
library(tidyr)
library(zoo)
library(skimr)

# need this for changes_one_country()
devtools::load_all(here::here("demspaces"))

# shortcut for plotting missing values when no date column is available
plotmiss <- function(x) {
  plot_missing(x, names(x), partial = "last")
}

# Minimum yearly threshold factor to use in the episode coder, see "min_f" in  
# ?episode_coder
MIN_F <- get_option("min_f")
```

```{r}
dv <- read_csv("../output/dv_data_1968_on.csv") %>%
  select(-country_name, -country_id, -country_text_id) %>%
  dplyr::filter(complete.cases(.)) %>%
  arrange(gwcode, year)

range(dv$year)
```


```{r code-dv}
cutpoints <- read_csv("../output/cutpoints.csv")
cp <- cutpoints[["up"]]
names(cp) <- cutpoints[["indicator"]]

dv_vars <- setdiff(names(dv), c("gwcode", "year"))

dv_piece <- list()
for (var_i in dv_vars) {
  dv_i <- dv
  dv_i <- dv %>%
    select(gwcode, year, !!var_i) %>%
    group_by(gwcode) %>%
    arrange(gwcode, year) %>%
    dplyr::mutate(
      # witches' magic that creates a 'dv_i' column that contains values as if in 
      # base R we did dv[, dv_i]; from now on i'll use dv_i for the remaining
      # computations
      var = !!rlang::sym(var_i),
      lag1_var = lag(var, 1L),
      y2y_diff_var = (var - lag1_var),
      # set y2y_diff to 0 if missing (first year of independence)
      y2y_diff_var = ifelse(is.na(y2y_diff_var), 0, y2y_diff_var)
    ) 
  
  # Bypass the normal flow to produce modified data
  # This produces the ERT-lite version of the outcome that was added in the 
  # spring 2022 update. 
  if (getOption("demspaces.version") >= "v12.1") {
    dv_i <- dv_i %>%
      mutate(var_change = changes_one_country(var, cp[[var_i]], type = "mod", 
                                              min_f = MIN_F),
             # actually the first 2 years of indy are missing with this algo,
             # but let's keep the same name
             var_change = ifelse(year < (min(year) + 2), 
                                 "first year of independence",
                                 var_change),
             # there is also an additional NA at the tail, set these to "same"
             # as with previous version of outcome coding
             var_change = ifelse(year==max(year) & is.na(var_change),
                                 "same",
                                 var_change)
             )
  } else {
    dv_i <- dv_i %>%
    mutate(var_change = case_when(
      y2y_diff_var > cp[[var_i]]  ~ "up",
      y2y_diff_var < -cp[[var_i]] ~ "down",
      year==min(year)      ~ "first year of independence",
      is.na(y2y_diff_var) & year!=min(year) ~ NA_character_,
      TRUE ~ "same"
    ))
  }
  # END of bypass
  
  dv_i <- dv_i %>%
    mutate(up = as.integer(var_change=="up"),
           lead1_up = lead(up, 1L),
           lead2_up = lead(up, 2L),
           next2_up = pmax(lead1_up, lead2_up),
           down = as.integer(var_change=="down"),
           lead1_down = lead(down, 1L),
           lead2_down = lead(down, 2L),
           next2_down = pmax(lead1_down, lead2_down)
    ) 
  # some states end during the data period; in these cases we can safely code
  # DV=0 since they are not really censored. 
  dv_i <- dv_i %>%
    ungroup() %>%
    mutate(
      next2_up = case_when(
        is.na(next2_up)   & (year < (max(year) - 1)) ~ 0L,
        TRUE ~ next2_up),
      next2_down = case_when(
        is.na(next2_down) & (year < (max(year) -1)) ~ 0L,
        TRUE ~ next2_down
      ))
  # rename variables
  dv_i <- dv_i %>%
    select(gwcode, year, var, lag1_var, y2y_diff_var, var_change, up, next2_up,
           down, next2_down) %>%
    setNames(c("gwcode", "year", paste0(var_i, c("", "_lag1", "_diff_y2y", 
                                                 "_change", "_up", "_up_next2", 
                                                 "_down", "_down_next2"))))
  
  dv_to_join <- select(dv_i, gwcode, year, !!var_i, 
                       ends_with("change"), ends_with("next2"), 
                       ends_with("up"), ends_with("down")) %>%
    # sub in the actual variable name, not the placeholders
    setNames(c("gwcode", "year", paste0("", names(.)[3:ncol(.)]))) %>%
    # prepend "dv" to vars we should use as IVs
    rename_at(vars(ends_with("change"), ends_with("next2")), ~ paste0("dv_", .))
  
  dv_piece[[var_i]] <- dv_to_join
}
```

Write the pieces to CSV files so we can quickly see on git when anything changes.

```{r}
for (i in seq_along(dv_piece)) {
  var_i <- names(dv_piece)[i]
  dv_to_join <- dv_piece[[var_i]]
  dv_to_join <- dv_to_join %>% select(-ends_with("_up"), -ends_with("_down"),
                                      -starts_with("v2"))
  fn <- paste0(var_i, ".csv")
  write_csv(dv_to_join, file.path("../output/dv-lists", fn))
}
```

Combine into one large DV set. 

```{r}
dv_data <- Reduce(left_join, x = dv_piece)
```


Add squared transformations for space indicators. Changes seem to be more common at middle values and less common in high and low (partly because beyond some range a change is impossible). 

```{r squared, eval=FALSE}
data("spaces")
stopifnot(all(spaces$Indicator %in% names(dv_data)))
dv_data = dv_data %>%
  mutate_at(vars(one_of(spaces$Indicator)), list(squared = ~`^`(., 2)))
```

Add markers for when up/down changes are impossible given the current indicator value and cutpoints (see [#15](https://github.com/vdeminstitute/demspaces/issues/15)). 

```{r no-change-possible-markers, eval=FALSE}
data("spaces")

for (ind in spaces$Indicator) {
  for (dir in c("up", "down")) {
    marker <- paste0("no_", dir, "_", ind)
    cp <- cutpoints[cutpoints$indicator==ind, ][[dir]]
    if (dir=="up") {
      dv_data[[marker]] <- as.integer((dv_data[[ind]]) > (1 - cp))
    } else {
      dv_data[[marker]] <- as.integer((dv_data[[ind]]) < cp)
    }
    # for debug/check, print these crosstabs
    # almost all column 1's should be in the 0 row, see #15
    #cat(paste0(ind, " ", dir), "\n")
    #print(table(dv_data[[marker]], dv_data[[paste0("dv_", ind, "_", dir, "_next2")]]))
  }
}
```

Take out the plain up and down dv versions.

```{r}
dv_data <- dv_data %>%
  select(-ends_with("up"), -ends_with("down"))

range(dv_data$year)

write_rds(dv_data, "../output/dv-data.rds")

# Record some stats on the DV data so that during updates it is easier to
# quickly tell whether the new data make sense
dv_data %>%
  ungroup() %>%
  select(ends_with("_change")) %>%
  tidyr::pivot_longer(everything(), names_to = "space") %>%
  count(space, value) %>%
  mutate(value = factor(value, levels = c("first year of independence", "same",
                                          "up", "down"))) %>%
  arrange(value) %>%
  tidyr::pivot_wider(names_from = "value", values_from = "n") %>%
  write_csv("../output/dv-summary.csv")
```

Check the values for one outcome/country to make sure they make sense:

```{r}
dv_data %>%
  dplyr::filter(gwcode==310) %>%
  select(gwcode, year, contains("veracc")) %>%
  dplyr::filter(year > 2010)
```


```{r}
skim(dv_data)
```

