---
title: "Population"
output: 
  github_document:
    toc: yes
---

*Last updated on: `r Sys.Date()`*

This file combines UN, WDI, and KSG population data to create a complete coverage dataset for 1950 to 2019, with only Abkhazia, South Ossetia, and Zanzibar missing. 

The data sources are:

- https://population.un.org/wpp/DataQuery/: select total population by sex and SDG regions; only 35 years can be exported at a time so three files are needed
- http://ksgleditsch.com/data/exppop.txt
- World Bank, via the WDI package

To update the data:

1. Download a new version of the UN population dataset at https://population.un.org/wpp/DataQuery/. Only 35 years can be exported at a time so three files are needed. Select total population by sex, "By Sex", "SDG subregions", on the popup yes, select years needed, "Both sexes combined", "By Years/Periods", "Export to Excel"; rename the output file to match the pattern in the `input/` folder. 
2. Delete "input/wdipop.csv"; it'll be downloaded and cached again. 
3. Search for "UPDATE:" for lines in this document that need attention, possibly.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(states)
library(WDI)
library(lubridate)
library(readxl)
library(kableExtra)
library(imputeTS)

wdi_add_gwcode <- function(x) {
  starty <- min(x$year)
  endy <- max(x$year)
  cy <- states::state_panel(starty, endy, useGW = TRUE, partial = "any")
  
  x <- x %>%
    mutate(gwcode = suppressWarnings(countrycode::countrycode(x$iso2c, "iso2c", "cown")),
           gwcode = as.integer(gwcode)) %>%
    mutate(gwcode = case_when(
      iso2c=="RS" ~ 340L,
      iso2c=="XK" ~ 347L,
      iso2c=="VN" ~ 816L,
      
      gwcode==255 ~ 260L,
      gwcode==679 ~ 678L,
      
      gwcode==970 ~ 971L,
      gwcode==946 ~ 970L,
      gwcode==947 ~ 973L,
      gwcode==955 ~ 972L,
      TRUE ~ gwcode
    ))
  
  # fix Czechoslovakia
  x$gwcode[x$gwcode==316 & x$year <= 1992] <- 315L
  
  # drop countries that unify
  x <- x %>%
    # pre-95 Serbia
    filter(!(gwcode==340 & year < 1995)) %>%
    filter(!(gwcode==260 & year < 1990)) %>%
    filter(!(gwcode==678 & year < 1990)) %>%
    # pre-75 Vietnam
    filter(!(gwcode==816 & year < 1975)) 
  
  x <- dplyr::select(x, -iso2c, -country)
  x <- x %>% filter(!is.na(gwcode))
  
  cy <- dplyr::left_join(cy, x, by = c("gwcode", "year"))
  cy
}

data(gwstates)
cnames <- gwstates %>%
  group_by(gwcode) %>% 
  slice(n()) %>%
  select(gwcode, country_name) %>%
  ungroup()
```

## Expanded population data from KSG

Data are at http://ksgleditsch.com/data/exppop.txt

```{r}
if (!file.exists("input/exppop.tsv")) {
  url <- "http://ksgleditsch.com/data/exppop.txt"
  download.file(url, destfile = "input/exppop.tsv")
}

ksg <- read_tsv("input/exppop.tsv")
```

The data range from `r min(ksg$year)` to `r max(ksg$year)`. 

```{r}
plot_missing(ksg, x = "pop", ccode = "idnum", time = "year", statelist = "GW")
```

The few tailing non-independent cases are alright, they were just independent for part of the year, not through Dec. 31. 

These data only go to 2004. Need to splice in updates. 

## WDI pop data

Via the WB API thanks to WDI package. 

```{r get-wdi}
# UPDATE: delete input/wdipop.csv to re-download
if (!file.exists("input/wdipop.csv")) {
  wdi_raw <- WDI(country = "all", indicator = "SP.POP.TOTL", 
               start = 1960, end = year(Sys.Date()), extra = FALSE) 
  write.csv(wdi_raw, "input/wdipop.csv", row.names = FALSE)
}

wdi_raw <- read.csv("input/wdipop.csv")

wdi <- wdi_raw %>% wdi_add_gwcode(.) %>%
  rename(pop = SP.POP.TOTL) %>%
  # change to pop in 1,000s
  mutate(pop = pop / 1e3)

plot_missing(wdi, x = "pop", ccode = "gwcode", time = "year", statelist = "GW")
```

## UN pop data

Splice together and clean the raw UN pop files. 

```{r read-un}
files <- dir("input", pattern = "UNPop", full.names = TRUE)

# This is the UN pop data for countries, taking out the region aggregations
un_raw <- files %>%
  # read in data
  map(., readxl::read_xlsx, sheet = 2, skip = 1) %>%
  # cut fat from the tibbles
  map(rename, iso3n = `ISO 3166-1 numeric code`) %>%
  map(dplyr::select, -Sex, -Note) %>%
  # now we can join them together
  purrr::reduce(full_join, by = c("iso3n", "Location")) %>%
  filter(iso3n < 900) %>%
  # change from wide to long
  gather(year, pop, -iso3n, -Location) %>%
  mutate(year = as.integer(year))

# UPDATE: drop the 2021 data; i just didn't want to export a file with a single
# year
un_raw <- un_raw[!un_raw$year==2021, ] 

# Add GW codes
un <- un_raw %>%
  mutate(gwcode = countrycode::countrycode(iso3n, "iso3n", "cown", warn = FALSE),
         gwcode = as.integer(gwcode)) %>%
  mutate(gwcode = case_when(
      Location=="Serbia" ~ 340L, 
      gwcode==255 ~ 260L,
      gwcode==679 ~ 678L,
      gwcode==970 ~ 971L,
      gwcode==946 ~ 970L,
      gwcode==947 ~ 973L,
      gwcode==955 ~ 972L,
      gwcode==817 ~ 816L,
      TRUE ~ gwcode
  )) 

un <- un %>%
  filter(!is.na(gwcode))
```

### Reconstruct some countries that later split

Since the UN data has series for all current countries going back to 1950, we can use those to re-construct countries like USSR and Yugoslavia that today are several countries.

#### Czechoslovakia

Czechia and Slovakia split on 1 January 1993. 

```{r}
add <- tibble(
  gwcode = 315, 
  year = 1950:1992,
  pop = rowSums(cbind(un$pop[un$gwcode==316 & un$year < 1993], 
                         un$pop[un$gwcode==317 & un$year < 1993]))
)
un <- bind_rows(un, add) %>%
  filter(!(gwcode==316 & year < 1993),
         !(gwcode==317 & year < 1993))
```

#### Pakistan pre-1971

Before East Pakistan became Bangladesh. 

```{r}
pak70 <- rowSums(cbind(un$pop[un$gwcode==770 & un$year < 1970], 
                       un$pop[un$gwcode==771 & un$year < 1970]))
un$pop[un$gwcode==770 & un$year < 1970] <- pak70
```

#### Yugoslavia/Serbia & Montenegro/Serbia

- 1990 and before: Slovenia, Croatia, BiH, Serbia, Kosovo, (UN does not seem to treat it separately), Montenegro, Macedonia
- 1991: Slovenia, Croatia, BiH, Serbia, Kosovo, (UN does not seem to treat it separately), Montenegro
- 1992 and until 2006: Serbia, Kosovo, Montenegro

```{r}
yugo <- tibble(
  gwcode = 345,
  year = 1950:2006,
  pop  = rowSums(cbind(
    # Serbia and Montenegro
    un$pop[un$gwcode==340 & un$year <= 2006],
    un$pop[un$gwcode==341 & un$year <= 2006],
    # Slovenia, Croatia, BiH
    c(un$pop[un$gwcode==349 & un$year <= 1991], rep(0, 15)),
    c(un$pop[un$gwcode==344 & un$year <= 1991], rep(0, 15)),
    c(un$pop[un$gwcode==346 & un$year <= 1991], rep(0, 15)),
    # Macedonia
    c(un$pop[un$gwcode==343 & un$year <= 1990], rep(0, 16))
  )))

plot(yugo$year, yugo$pop, ylim = c(0, 25e3), type = "l")

un <- un %>%
  filter(!(gwcode==340 & year < 2006),
         !(gwcode==341 & year < 2006),
         !(gwcode==343 & year < 1991),
         !(gwcode==344 & year < 1992),
         !(gwcode==346 & year < 1992),
         !(gwcode==349 & year < 1992)) %>%
  bind_rows(., yugo) 
```

#### USSR/Russia

- 1990 and before: Russia (365), the Baltics (366, 367, 368), Ukraine (369), Belarus (370), Armenia (371), Georgia (372), Azerbaijan (373), central Asia (701, 702, 703, 704, 705) 
- 1991 and on: 365

```{r}
ussr <- tibble(
  gwcode = 365,
  year = 1950:1990,
  pop = rowSums(cbind(
    un$pop[un$gwcode==365 & un$year < 1991],
    un$pop[un$gwcode==366 & un$year < 1991],
    un$pop[un$gwcode==367 & un$year < 1991],
    un$pop[un$gwcode==368 & un$year < 1991],
    un$pop[un$gwcode==369 & un$year < 1991],
    un$pop[un$gwcode==370 & un$year < 1991],
    un$pop[un$gwcode==371 & un$year < 1991],
    un$pop[un$gwcode==372 & un$year < 1991],
    un$pop[un$gwcode==373 & un$year < 1991],
    un$pop[un$gwcode==701 & un$year < 1991],
    un$pop[un$gwcode==702 & un$year < 1991],
    un$pop[un$gwcode==703 & un$year < 1991],
    un$pop[un$gwcode==704 & un$year < 1991],
    un$pop[un$gwcode==705 & un$year < 1991]
  ))
)

un <- un %>% 
  filter(
    !(gwcode==365 & year < 1991),
    !(gwcode==366 & year < 1991),
    !(gwcode==367 & year < 1991),
    !(gwcode==368 & year < 1991),
    !(gwcode==369 & year < 1991),
    !(gwcode==370 & year < 1991),
    !(gwcode==371 & year < 1991),
    !(gwcode==372 & year < 1991),
    !(gwcode==373 & year < 1991),
    !(gwcode==701 & year < 1991),
    !(gwcode==702 & year < 1991),
    !(gwcode==703 & year < 1991),
    !(gwcode==704 & year < 1991),
    !(gwcode==705 & year < 1991)
    ) %>%
  bind_rows(., ussr)
```


#### Sudan/South Sudan

South Sudan independent on `r sfind("South Sudan", "G&W")$start`. So use combined for 2011 and before. 

```{r}
sud <- rowSums(cbind(un$pop[un$gwcode==625 & un$year <= 2011], 
                     un$pop[un$gwcode==626 & un$year <= 2011]))
un$pop[un$gwcode==625 & un$year <= 2011] <- sud
```

#### Indonesia/East Timor

East Timor gained independence in 2002, so use combined for 2001 and before. 

```{r}
ind <- rowSums(cbind(un$pop[un$gwcode==850 & un$year <= 2001], 
                     un$pop[un$gwcode==860 & un$year <= 2001]))
un$pop[un$gwcode==850 & un$year <= 2001] <- ind
```

#### Check left-over discrepancies

Unions like GDR joining FRG are problematic and have to be fixed outside. 

- German re-unification
- Yemeni unification
- Vietnamese unification

Other discrepancies in the data:

```{r}
# UPDATE: end_year
end_year <- 2020
gw <- state_panel(1950, end_year, partial = "any")
gw_not_in_un <- gw %>%
  anti_join(un, by = c("gwcode", "year")) %>%
  group_by(gwcode) %>%
  mutate(seq = id_date_sequence(year, "year")) %>%
  group_by(gwcode, seq) %>%
  summarize(years = paste0(range(year), collapse = " - "),
            .groups = "drop") %>%
  left_join(cnames, by = "gwcode") 
gw_not_in_un %>%
  knitr::kable(caption = "GW CYs not in UN") 
un_not_in_gw <- un %>%
  anti_join(gw, by = c("gwcode", "year")) %>%
  group_by(gwcode) %>%
  mutate(seq = id_date_sequence(year, "year")) %>%
  group_by(gwcode, seq) %>%
  summarize(years = paste0(range(year), collapse = " - "),
            .groups = "drop") %>%
  left_join(cnames, by = "gwcode") 
un_not_in_gw %>%
  knitr::kable(caption = "UN CYs not in GW")


plot_missing(un, x = "pop", ccode = "gwcode", time = "year", statelist = "GW")
```


## Combined and overlap

```{r}
ksg2 <- ksg %>%
  rename(gwcode = idnum, pop_ksg = pop, source_ksg = source) %>%
  select(gwcode, year, pop_ksg, source_ksg)
wdi2 <- wdi %>%
  rename(pop_wdi = pop) %>%
  select(gwcode, year, pop_wdi)
un2 <- un %>%
  rename(pop_un = pop) %>%
  select(gwcode, year, pop_un)
joint <- list(ksg2, wdi2, un2) %>%
  purrr::reduce(full_join, by = c("gwcode", "year")) %>%
  tidyr::gather(source, pop, -gwcode, -year, -source_ksg) %>%
  # add an indicator for whether the sources have overlapping coverage
  group_by(gwcode, year) %>%
  mutate(overlap = !any(is.na(pop))) %>%
  ungroup()

joint_wide <- joint %>%
  spread(source, pop)
```

The next plot shows the UN, WDI, and KSG population series for each country. 

```{r}
ggplot(joint, aes(x = year, y = pop, group = interaction(gwcode, source),
                  color = source)) +
  geom_line() +
  scale_y_log10() +
  theme_minimal()
```

There are quite some divergences. The UN and WDI data seem to generally be much more smooth than the KSG data, which has sometimes drastic shifts. I would say the UN values would be preferable where possible, they are more smooth than WDI and are easier to use to reconstruct historical state unions like Yugoslavia that are now several states. 

Try to sample some series to get a better look.

```{r}
countries <- c(2, 260, 344, 345, 365, 436, 540, 645, 651, 678, 710)
joint %>%
  filter(year > 1959) %>%
  filter(gwcode %in% countries) %>%
  left_join(cnames, by = c("gwcode")) %>%
  ggplot(aes(x = year, y = pop, color = source)) +
  facet_wrap(~ country_name, scales = "free_y") +
  geom_line() + 
  theme_minimal()
```

So it seems that UN is generally preferable, except for cases like Germany before 1990, where historical data were adjusted in WDI to ignore country changes. There it would be preferable to use KSG. 

### Within country covariances between UN and KSG

Check the country correlations.

```{r, fig.height=4}
cors <- joint %>%
  spread(source, pop) %>%
  group_by(gwcode) %>%
  summarize(cor = tryCatch(
    cor(pop_ksg, pop_un, use = "complete.obs"),
    error = function(e) NA_real_))

ggplot(cors, aes(x = cor)) +
  geom_histogram(binwidth = 0.1) +
  theme_minimal()
```

Most of them are 0.9 or higher. What about the exceptions?

```{r, fig.height=7}
countries <- filter(cors, cor < 0.85) %>% pull(gwcode)
joint %>%
  filter(year > 1959) %>%
  filter(gwcode %in% countries) %>%
  left_join(cnames, by = c("gwcode")) %>%
  ggplot(aes(x = year, y = pop, color = source)) +
  facet_wrap(~ country_name, scales = "free_y") +
  geom_line() + 
  theme_minimal()
```

UN seems fine on these except Germany pre-1990, USSR pre-1990, and Pakistan pre-1971. 

### Can we combine KSG for pre-50 with UN?

Check to see how well they are aligned. 

```{r}
# Only look at countries where GW is meeting UN; UN has too many
weld_countries <- joint %>% 
  filter(year==1949 & source == "pop_ksg") %>%
  pull(gwcode)
weld <- joint %>%
  filter(year > 1945 & year < 1956 & gwcode %in% weld_countries) %>%
  left_join(cnames, by = c("gwcode")) %>%
  filter(source!="pop_wdi") %>%
  spread(source, pop) %>%
  mutate(diff = (pop_ksg - pop_un) / (.5*(pop_ksg+pop_un)))

# Countries with divergence in meeting up year
lookat <- filter(weld, year==1950 & abs(diff) > .05) %>% pull(gwcode)
  
weld %>%
  gather(source, pop, pop_ksg, pop_un) %>%
  filter(gwcode %in% lookat) %>%
  ggplot(., aes(x = year, y = pop, color = source, group = interaction(source, gwcode))) +
  geom_line() + 
  theme_minimal() +
  scale_y_log10()
```

Some divergences greater than 5%, but let's gloss over those for now. 

## Start imputing/combining

Take UN as the preferable source, but drop in KSG for known deviations and pre-1950. For Kosovo we use WDI. 

```{r}
master <- state_panel(1816, max(joint$year), partial = "any") %>%
  mutate(gw = TRUE)
pop <- joint_wide %>%
  mutate(
    pop = case_when(
      year > 1949 ~ pop_un,
      year < 1950 ~ pop_ksg,
      TRUE ~ NA_real_),
    source = case_when(
      year > 1949 ~ "un",
      year < 1950 ~ "ksg",
      TRUE ~ NA_character_)
  ) %>%
  right_join(., master, by = c("gwcode", "year")) %>%
  replace_na(list(gw = FALSE))
```


### Germany

```{r}
joint_wide %>%
  filter(gwcode==265 & year > 1985 & year < 1995)

## use 1990 and before KSG for Germany
idx <- pop$gwcode==265 & pop$year <= 1990
pop$pop[idx] <- pop$pop_ksg[idx]
pop$source[idx] <- "ksg"
```

### Vietnam

```{r}
joint_wide %>%
  filter(gwcode==816 & year > 1970 & year < 1980)

## use 1974 and before KSG for DRV
idx <- pop$gwcode==816 & pop$year <= 1974
pop$pop[idx] <- pop$pop_ksg[idx]
pop$source[idx] <- "ksg"
```

```{r}
joint_wide %>%
  filter(gwcode==817 & year > 1970 & year < 1980)

## use 1975 and before KSG for RV
idx <- pop$gwcode==817 & pop$year <= 1975
pop$pop[idx] <- pop$pop_ksg[idx]
pop$source[idx] <- "ksg"
```

### Yemen


```{r}
joint_wide %>%
  filter(gwcode==678 & year > 1985 & year < 1995)

## use 1989 and before KSG for north Yemen
idx <- pop$gwcode==678 & pop$year <= 1989
pop$pop[idx] <- pop$pop_ksg[idx]
pop$source[idx] <- "ksg"
```

```{r}
joint_wide %>%
  filter(gwcode==680 & year > 1985 & year < 1995)

## use KSG for south Yemen
idx <- pop$gwcode==680 & pop$year <= 1990
pop$pop[idx] <- pop$pop_ksg[idx]
pop$source[idx] <- "ksg"
```

### Kosovo

```{r}
joint_wide %>%
  filter(gwcode==347 & year > 2006 & year < 2015)

# Get Kosovo series and kalman smooth
idx <- pop$gwcode==347
kos <- pop$pop_wdi[idx]
kos <- imputeTS::na_kalman(kos, "auto.arima")
pop$pop[idx] <- kos
pop$source[idx] <- "wdi"
```

### Tibet

Tibet is missing in UN data, so use KSG as with 1949 and before. 

```{r tibet}
joint_wide %>%
  filter(gwcode==711 & year > 1945 & year < 1955)

# Use KSG for 1950 as well
idx <- pop$gwcode==711 & pop$year == 1950
pop$pop[idx] <- pop$pop_ksg[idx]
pop$source[idx] <- "ksg"
```

### Czechoslovakia

Missing first year (1918), backwards impute.

```{r czechoslovakia}
joint_wide %>%
  filter(gwcode==315 & year < 1921)

pop %>% 
  filter(gwcode==315 & year < 1921)
```

Plot the pop series:

```{r plot-czechoslovakia}
idx <- pop$gwcode==315
csk <- pop$pop_ksg[idx]
plot(pop$year[idx], csk)
```

It is quite jumpy, so use only pre-1937

```{r}
idx2 <- pop$gwcode==315 & pop$year <= 1937
csk  <- pop$pop_ksg[idx2]
csk  <- rev(imputeTS::na.kalman(rev(csk), "auto.arima"))
pop$pop[idx2]   <- csk
pop$source[idx2] <- "ksg"

# Verify via plot
idx <- pop$gwcode==315
csk <- pop$pop[idx]
plot(pop$year[idx], csk)
```

## Get ready to write final data

### Check values for splitting/joining countries

```{r}
countries <- c(260, 265, 
               345, 340, 341, 343, 344, 346, 347, 349, 
               816, 817,
               678, 680,
               365, 366, 367, 368)
pop %>%
  filter(gwcode %in% countries) %>%
  left_join(cnames, by = c("gwcode")) %>%
  ggplot(aes(x = year, y = pop)) +
  facet_wrap(~ country_name, scales = "free") +
  geom_line() + 
  theme_minimal()
```

### Missingness

```{r}
plot_missing(pop, x = "pop", ccode = "gwcode", time = "year", statelist = "GW")
```

The final data is complete for 1950 to 2019, except for Abkhazia, South Ossetia, and Zanzibar. 

```{r}
pop %>%
  filter(is.na(pop)) %>%
  group_by(gwcode) %>%
  summarize(years = paste0(range(year), collapse = " - "), N = n()) %>%
  mutate(country_name = country_names(gwcode)) %>%
  knitr::kable()
```


```{r}
pop %>%
  select(gwcode, year, pop) %>%
  write_csv(., path = "output/population.csv")
```

