---
title: "GDP imputation"
author: "Andreas Beger, Predictive Heuristics"
date: "`r Sys.Date()`"
output: 
  github_document: 
    toc: yes
---

*Last updated on: `r Sys.Date()`*

For updating the data, all items requiring attention are marked with "UPDATE:".

UPDATE: This script uses `population.csv` as an input for calculating GDP per capita. Drop in an updated version from `population/` if needed. 

The WDI GDP data, cached in the input folder, also need to be updated. Delete `input/wdigdp.csv` for that to happen automatically (or see the UPDATE: below).

The other inputs (KSG and UN GDP) only concern historical data so they should not need to be updated. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(WDI)
library(states)
library(lubridate)
library(countrycode)
library(lme4)
library(imputeTS)

gdp_wdi_add_gwcode <- function(x) {
  starty <- min(x$year)
  endy <- max(x$year)
  cy <- states::state_panel(starty, endy, partial = "any", useGW = TRUE)

  x <- x %>%
    mutate(gwcode = countrycode::countrycode(x$iso2c, "iso2c", "cown", warn = FALSE),
           gwcode = as.integer(gwcode)) %>%
    mutate(gwcode = case_when(
      iso2c=="RS" ~ 340L,
      iso2c=="XK" ~ 347L,

      gwcode==255 ~ 260L,
      gwcode==679 ~ 678L,

      gwcode==970 ~ 971L,
      gwcode==946 ~ 970L,
      gwcode==947 ~ 973L,
      gwcode==955 ~ 972L,
      TRUE ~ gwcode
    ))

  # drop countries that unify
  x <- x %>%
    filter(!(gwcode==260 & year < 1990)) %>%
    filter(!(gwcode==678 & year < 1990))

  x <- dplyr::select(x, -iso2c, -country)
  x <- x %>% filter(!is.na(gwcode))

  cy <- dplyr::left_join(cy, x, by = c("gwcode", "year"))
  cy
}

gdp_un_add_gwcode <- function(x) {
  starty <- min(x$year)
  endy <- max(x$year)
  cy <- states::state_panel(starty, endy, partial = "any", useGW = TRUE)

  x <- x %>%
    mutate(cowcode = countrycode(country_name, "country.name", "cown", warn = FALSE),
           cowcode = as.integer(cowcode))

  x <- x %>%
    mutate(gwcode = case_when(
      country_name=="Democratic Republic of Vietnam" ~ 816L,
      # UN seems to have correct series for Yugo/Serbia
      country_name=="Serbia" & year > 2006 ~ 340L,
      country_name=="Serbia" & year <= 2006 ~ 345L,
      # need to add 2006 for Serbia
      cowcode==255 ~ 260L,
      cowcode==679 ~ 678L,
      cowcode==955 ~ 972L,
      # Czechoslovakia/CR also treated correctly, and they had a nice split date
      cowcode==316 & year < 1993 ~ 315L,
      TRUE ~ cowcode
    ))
  serbia2006 <- x[x$country_name=="Serbia" & x$year==2006, ]
  serbia2006$gwcode <- 340L

  x <- bind_rows(x, serbia2006) %>%
    arrange(gwcode, year)

  # drop countries that unify, historic GDP values are incorrect
  x <- x %>%
    filter(!(gwcode==260 & year < 1990)) %>%
    filter(!(gwcode==678 & year < 1990)) %>%
    filter(!(gwcode==816 & year < 1975))

  in_un_not_in_master <- anti_join(x, cy, by = c("year", "gwcode")) %>%
    group_by(country_name) %>%
    summarize(gwcode = unique(gwcode)[1], year = paste0(range(year), collapse = " - "))

  in_master_not_in_un <- anti_join(cy, x, by = c("gwcode", "year")) %>%
    group_by(gwcode) %>%
    summarize(year = paste0(range(year), collapse = " - "))

  x <- dplyr::select(x, -country_name, -country_id, -cowcode)
  x <- x %>% filter(!is.na(gwcode))

  cy <- dplyr::left_join(cy, x, by = c("gwcode", "year"))
  cy
}

load_inputs <- function() {
  # don't want column specification messages
  suppressMessages({
    wdigdp <- read.csv("input/wdigdp.csv")[, c("iso2c", "country", "year", "NY.GDP.MKTP.KD")]
    
    ksggdp <- read_delim("input/expgdpv6.0/gdpv6.txt", delim = "\t") %>%
      rename(gwcode = statenum) %>%
      select(-stateid)

    ungdp <- read_csv("input/UNgdpData.csv") %>%
      select(country_name, country_id, year, gdp_2010USD)
    
    pop <- read_csv("input/population.csv")
  })

  list(wdigdp = wdigdp, ksggdp = ksggdp, ungdp = ungdp, pop = pop)
}

# This is a top-level function for cleaning the GDP data
# impute: TRUE/FALSE impute missing series values?
# inputs: a 3-item list with WDI GDP data, KSG GDP, UN GDP, and pop data
#         (names: c("wdigdp", "ksggdp", "ungdp", "pop"))
gdp_get_yearly <- function(impute, inputs = load_inputs()) {

  wdigdp <- inputs$wdigdp
  ksggdp <- inputs$ksggdp
  ungdp  <- inputs$ungdp
  pop <- inputs$pop

  # if the latest year (or any others) are completely missing, drop them
  dropyear <- wdigdp %>%
    group_by(year) %>%
    summarize(missing = sum(is.na(NY.GDP.MKTP.KD))==n(),
              .groups = "drop") %>%
    filter(missing) %>%
    pull(year)
  if (length(dropyear) > 0) {
    cat("Year(s) ", paste0(dropyear, collapse = ", "), "are complete missing, droping\n")
    wdigdp <- wdigdp %>%
      filter(!year %in% dropyear)
  }
  
  # add standard country codes
  wdi <- gdp_wdi_add_gwcode(wdigdp)
  ungdp <- gdp_un_add_gwcode(ungdp)

  # Combine
  joint <- wdi %>%
    full_join(., ksggdp, by = c("gwcode", "year")) %>%
    select(-pop, -rgdppc, -cgdppc) %>%
    mutate(realgdp = realgdp*1e6) %>%
    full_join(., ungdp, by = c("gwcode", "year")) %>%
    arrange(gwcode, year)

  mdl_un <- lm(NY.GDP.MKTP.KD ~ -1 + gdp_2010USD, data = joint)
  joint <- joint %>%
    mutate(un_gdp.rescaled = predict(mdl_un, newdata = joint))

  mdl_ksg <- lmer(log(NY.GDP.MKTP.KD) ~ -1 + log(realgdp) + (log(realgdp)|gwcode), data = joint)
  joint <- joint %>%
    mutate(ksg_gdp.rescaled = exp(predict(mdl_ksg, newdata = joint, allow.new.levels = TRUE)))

  joint <- joint %>%
    mutate(NY.GDP.MKTP.KD = case_when(
      # special treatment for Qatar 1971, where otherwise a big jump occurs
      gwcode==694 & year==1971 ~ filter(joint, gwcode==694) %>% 
        pull(un_gdp.rescaled) %>% rev() %>% imputeTS::na_kalman() %>% tail(1),
      is.na(NY.GDP.MKTP.KD) & !is.na(un_gdp.rescaled) ~ un_gdp.rescaled,
      is.na(NY.GDP.MKTP.KD) & !is.na(ksg_gdp.rescaled) ~ ksg_gdp.rescaled,
      TRUE ~ NY.GDP.MKTP.KD
    ))

  joint <- joint %>%
    # take out components; don't need anymore
    select(gwcode, year, NY.GDP.MKTP.KD) %>%
    # add GDP growth
    arrange(gwcode, year) %>%
    group_by(gwcode) %>%
    mutate(
      NY.GDP.MKTP.KD.ZG = (NY.GDP.MKTP.KD - lag(NY.GDP.MKTP.KD)) / 
        lag(NY.GDP.MKTP.KD) * 100) %>%
    ungroup()

  if (impute) {
    # forward and backward impute GDP
    joint <- joint %>%
      group_by(gwcode) %>%
      mutate(NY.GDP.MKTP.KD = imputeTS::na_kalman(NY.GDP.MKTP.KD)) %>%
      ungroup()
    joint <- joint %>%
      group_by(gwcode) %>%
      arrange(desc(year)) %>%
      mutate(NY.GDP.MKTP.KD = imputeTS::na_kalman(NY.GDP.MKTP.KD)) %>%
      ungroup()

    # update GDP change calculation; use backward impute for first year
    joint <- joint %>%
      group_by(gwcode) %>%
      arrange(year) %>%
      mutate(NY.GDP.MKTP.KD.ZG = (NY.GDP.MKTP.KD - lag(NY.GDP.MKTP.KD)) / lag(NY.GDP.MKTP.KD) * 100) %>%
      arrange(desc(year)) %>%
      mutate(NY.GDP.MKTP.KD.ZG = tryCatch({
        suppressWarnings(imputeTS::na_kalman(NY.GDP.MKTP.KD.ZG, model = "auto.arima"))
      }, error = function(e) NY.GDP.MKTP.KD.ZG)) %>%
      arrange(year) %>%
      ungroup()
  }

  # Add GDP per capita
  joint <- joint %>%
    left_join(pop, by = c("gwcode", "year")) %>%
    group_by(gwcode) %>%
    arrange(year) %>%
    mutate(NY.GDP.PCAP.KD = NY.GDP.MKTP.KD / (pop*1e3),
           NY.GDP.PCAP.KD.ZG = (NY.GDP.PCAP.KD - lag(NY.GDP.PCAP.KD)) / lag(NY.GDP.PCAP.KD) * 100) %>%
    select(-pop) %>%
    # backwards impute first year of GDP per capita change
    group_by(gwcode) %>%
    arrange(desc(year)) %>%
    mutate(NY.GDP.PCAP.KD.ZG = tryCatch({
      suppressWarnings(imputeTS::na_kalman(NY.GDP.PCAP.KD.ZG, model = "auto.arima"))
    }, error = function(e) NY.GDP.PCAP.KD.ZG)) %>%
    arrange(year) %>%
    ungroup()

  joint$year <- as.integer(joint$year)

  joint
}


```


## WDI GDP data

Relevant WDI indicators:

```
"NY.GDP.PCAP.PP.KD.ZG"
"NY.GDP.PCAP.PP.KD"
"NY.GDP.PCAP.KD.ZG"
"NY.GDP.PCAP.KD"
"NY.GDP.MKTP.PP.CD"  # GDP per capita (constant 2011 international $)
"NY.GDP.MKTP.PP.KD"  # GDP per capita (constant 2020 US$)
"NY.GDP.MKTP.KD"     # GDP (constant 2010 US$)
"NY.GDP.MKTP.KD.ZG"  # GDP growth
"SP.POP.TOTL"
```

UPDATE: delete `input/wdigdp.csv` and run the chunk below to re-download and cache the latest WDI GDP data. 

```{r cache-wdi}
if (!file.exists("input/wdigdp.csv")) {
  wdi1 <- WDI(country = "all", start = 1960, end = 2020,
              indicator = c("NY.GDP.MKTP.PP.KD"))
  wdi2 <- WDI(country = "all", start = 1960, end = 2020,
              indicator = c("NY.GDP.MKTP.PP.CD"))
  wdi3 <- WDI(country = "all", start = 1960, end = 2020,
              indicator = c("NY.GDP.MKTP.KD"))
  wdigdp <- Reduce(left_join, list(wdi1, wdi2, wdi3))
  write.csv(wdigdp, file = "input/wdigdp.csv", row.names = FALSE)
}

wdigdp <- read.csv("input/wdigdp.csv")
```


```{r}
wdi <- gdp_wdi_add_gwcode(wdigdp)

plot_missing(wdi, "NY.GDP.MKTP.KD", "gwcode", time = "year", 
             statelist = "GW") +
  ggtitle("NY.GDP.MKTP.KD")
plot_missing(wdi, "NY.GDP.MKTP.PP.KD", "gwcode",  time =  "year", 
             statelist = "GW") +
  ggtitle("NY.GDP.MKTP.PP.KD")
plot_missing(wdi, "NY.GDP.MKTP.PP.CD", "gwcode",  time =  "year", 
             statelist = "GW") +
  ggtitle("NY.GDP.MKTP.PP.CD")

```


## KSG expanded GDP

```{r}
ksggdp <- read_delim("input/expgdpv6.0/gdpv6.txt", delim = "\t") %>%
  rename(gwcode = statenum) %>%
  select(-stateid)

plot_missing(ksggdp, "realgdp", "gwcode", time = "year", statelist = "GW")
```

## UN GDP data

```{r}
ungdp <- read_csv("input/UNgdpData.csv") %>%
  select(country_name, country_id, year, gdp_2010USD) 

ungdp <- gdp_un_add_gwcode(ungdp)  
plot_missing(ungdp, "gdp_2010USD", "gwcode", time = "year", statelist = "GW")
```

## Combine data

```{r}
joint <- wdi %>%
  full_join(., ksggdp, by = c("gwcode", "year")) %>%
  select(-pop, -rgdppc, -cgdppc) %>%
  mutate(realgdp = realgdp*1e6) %>%
  full_join(., ungdp, by = c("gwcode", "year")) 
  

# Example countries to look at below
countries <- unique(c(
  c(2, 200, 220, 260, 290, 315, 740, 710),
  sample(unique(joint$gwcode), 4)))
```

## Overlap between UN and WDI

The UN GDP data is almost completely correlated with WDI GDP.

```{r}
# the UN GDP is almost completely correlated with WDI GDP
sum(complete.cases(joint[, c("gdp_2010USD", "NY.GDP.MKTP.KD")]))
cor(joint$gdp_2010USD, joint$NY.GDP.MKTP.KD, use = "complete.obs")

plot(log10(joint$gdp_2010USD), log10(joint$NY.GDP.MKTP.KD))
```

Does it add any non-missing values? Yes, about 800 or so. 

```{r}
# does it add any non-missing values?
joint %>%
  mutate(un_gdp_missing = is.na(gdp_2010USD),
         wdi_gdp_missing = is.na(NY.GDP.MKTP.KD)) %>%
  group_by(un_gdp_missing, wdi_gdp_missing) %>%
  summarize(n = n())
```

For which countries? Somalia, Syria, ...

```{r}
# which countries?
adds <- joint %>% 
  filter(is.na(NY.GDP.MKTP.KD) & !is.na(gdp_2010USD)) %>%
  group_by(gwcode) %>%
  summarize(adds = n())
head(arrange(adds, desc(adds)))
```


```{r}
# look at some examples of those
set.seed(1343)
countries2 <- unique(c(c(290, 345), 
                       sample(adds$gwcode, 8)))

mdl <- lm(NY.GDP.MKTP.KD ~ -1 + gdp_2010USD, data = joint)
joint <- joint %>%
  mutate(gdp_2010USD.rescaled = predict(mdl, newdata = joint))
joint %>%
  gather(var, value, -gwcode, -year, -origin, -realgdp, -NY.GDP.MKTP.PP.KD, -gdp_2010USD) %>%
  filter(gwcode %in% countries2) %>%
  ggplot(aes(x = year, y = value, colour = var, group = interaction(gwcode, var))) +
  geom_line(alpha = .5) +
  facet_wrap(~ gwcode, scales = "free_y")
joint %>%
  gather(var, value, -gwcode, -year, -origin, -NY.GDP.MKTP.PP.KD, -gdp_2010USD) %>%
  filter(gwcode %in% countries) %>%
  ggplot(aes(x = year, y = value, colour = var, group = interaction(gwcode, var))) +
  geom_line(alpha = .5) +
  facet_wrap(~ gwcode, scales = "free_y")
```

Rescaled UN GDP matches WDI very well, it seems. Adjusted R^2 is basically 1 and so is the coefficient. 

## Overlap between KSG expanded and WDI

```{r}
sum(complete.cases(joint[, c("realgdp", "NY.GDP.MKTP.PP.KD")]))
cor(joint$realgdp, joint$NY.GDP.MKTP.PP.KD, use = "complete.obs")

sum(complete.cases(joint[, c("realgdp", "NY.GDP.MKTP.KD")]))
cor(joint$realgdp, joint$NY.GDP.MKTP.KD, use = "complete.obs")
```

Plain linear rescaling doesn't work well.

```{r}
# Plain linear rescaling; doesn't work well
mdl <- lm(NY.GDP.MKTP.KD ~ -1 + realgdp, data = joint)
summary(mdl)
joint <- joint %>%
  mutate(realgdp.rescaled = predict(mdl, newdata = joint))
joint %>%
  gather(var, value, -gwcode, -year, -origin, -NY.GDP.MKTP.PP.KD, -starts_with("gdp_2010")) %>%
  filter(gwcode %in% countries) %>%
  ggplot(aes(x = year, y = value, colour = var, group = interaction(gwcode, var))) +
  geom_line(alpha = .5) +
  facet_wrap(~ gwcode, scales = "free_y")

plot(log10(joint$realgdp), log10(joint$NY.GDP.MKTP.KD))
plot(log10(joint$realgdp.rescaled), log10(joint$NY.GDP.MKTP.KD))
```

Try log model, also doesn't work super well. 

```{r}
# Try log log model; also doesn't work well
mdl <- lm(log(NY.GDP.MKTP.KD) ~ -1 + log(realgdp), data = joint)
summary(mdl)
joint <- joint %>%
  mutate(realgdp.rescaled2 = exp(predict(mdl, newdata = joint)))
joint %>%
  gather(var, value, -gwcode, -year, -origin, -NY.GDP.MKTP.PP.KD, -starts_with("gdp_2010")) %>%
  filter(gwcode %in% countries) %>%
  ggplot(aes(x = year, y = value, colour = var, group = interaction(gwcode, var))) +
  geom_line() +
  facet_wrap(~ gwcode, scales = "free_y")

plot(log10(joint$realgdp), log10(joint$NY.GDP.MKTP.KD))
abline(a = 0, b = 1)
plot(log10(joint$realgdp.rescaled2), log10(joint$NY.GDP.MKTP.KD))
abline(a = 0, b = 1)
```

Try scaling by country:

```{r}
# try country-varying scaling factors; this works fairly well
library("lme4")
mdl <- lmer(log(NY.GDP.MKTP.KD) ~ -1 + log(realgdp) + (log(realgdp)|gwcode), data = joint)
joint <- joint %>%
  mutate(realgdp.rescaled3 = exp(predict(mdl, newdata = joint, allow.new.levels = TRUE)))
joint %>%
  gather(var, value, -gwcode, -year, -origin, -realgdp.rescaled, -realgdp.rescaled2,
         -starts_with("gdp_2010"), -NY.GDP.MKTP.PP.KD) %>%
  filter(gwcode %in% countries) %>%
  ggplot(aes(x = year, y = value, colour = var, group = interaction(gwcode, var))) +
  geom_line() +
  facet_wrap(~ gwcode, scales = "free_y")

plot(log10(joint$realgdp), log10(joint$NY.GDP.MKTP.KD))
abline(a = 0, b = 1)
plot(log10(joint$realgdp.rescaled3), log10(joint$NY.GDP.MKTP.KD))
abline(a = 0, b = 1)
```


## Joint model of UN and KSG predicting WDI

```{r}
mdl_combo <- lmer(log(NY.GDP.MKTP.KD) ~ -1 + log(gdp_2010USD) + log(realgdp) + (log(realgdp)|gwcode), data = joint)
joint <- joint %>%
  mutate(NY.GDP.MKTP.KD.hat = exp(predict(mdl_combo, newdata = joint, allow.new.levels = TRUE)))
joint %>%
  gather(var, value, -gwcode, -year, -origin, -realgdp, -realgdp.rescaled, -realgdp.rescaled2, -realgdp.rescaled3,
         -starts_with("gdp_2010"), -NY.GDP.MKTP.PP.KD) %>%
  filter(gwcode %in% countries) %>%
  ggplot(aes(x = year, y = value, colour = var, group = interaction(gwcode, var))) +
  geom_line() +
  facet_wrap(~ gwcode, scales = "free_y")
joint %>%
  gather(var, value, -gwcode, -year, -origin, -realgdp, -realgdp.rescaled, -realgdp.rescaled2, -realgdp.rescaled3,
         -starts_with("gdp_2010"), -NY.GDP.MKTP.PP.KD) %>%
  filter(gwcode %in% countries2) %>%
  ggplot(aes(x = year, y = value, colour = var, group = interaction(gwcode, var))) +
  geom_line() +
  facet_wrap(~ gwcode, scales = "free_y")
```

This works well, but cannot predict when either KSG or UN is missing, so not useful in practice for filling in WDI gaps.

## Conclusion

Four step imputation procedure:

1. Acquire the WDI data
2. Where WDI is missing, drop in UN GDP figures, scaled by a linear model.
3. Where WDI is missing, drop in KSG figures, scaled by a log-linear country-varying scaling model.
4. Model-based extrapolation: use Kalman-smoothing to forward extrapolate missing GDP values (most notably Taiwan and several countries missing current year GDP values) and backward extrapolate GDP growth in first year of existences of a country.  

Check leftover missing values before impute:

```{r}
joint <- gdp_get_yearly(impute = FALSE)

plot_missing(joint, "NY.GDP.MKTP.KD", "gwcode", time = "year", statelist = "GW")
  
still_missing <- joint %>% 
  filter(is.na(NY.GDP.MKTP.KD)) %>% 
  group_by(gwcode) %>%
  summarize(n = n(),
            years = paste0(range(year), collapse = " - ")) %>%
  arrange(desc(n))
still_missing
```

Use Kalman smoothing to extrapolate the leftover trailing missing values, and backwards extrapolate first year missing GDP growth. 

```{r}
joint <- gdp_get_yearly(impute = TRUE)

plot_missing(joint, "NY.GDP.MKTP.KD", "gwcode", time = "year", statelist = "GW")
```

## Check GDP per capita

This uses "population.csv" from the population module.

At least one of the combined GDP values--Qatar in 1971--is clunky in that there is a big discrepancy. This gives Qatar 1971 an inordinarily high GDP per capita value. Solved by backward imputing GDP instead of taking KSG value. 

```{r check-qatar}
check <- joint[joint$gwcode==694, ]
par(mfrow = c(2, 2))
for (vn in setdiff(names(check), c("gwcode", "year"))) {
  plot(check$year, check[[vn]], type = "l", main = vn, xlab = NA, ylab = NA)
}
```

I guess this is fine. (Qatar did really have these crazy high growth periods, checked the 2001 on data.)

## Done, record summary stats and save

```{r}
# Keep a summary of the data so changes in the future are easier to track on 
# git
df <- joint
stats <- list(
  Class           = paste0(class(df), collapse = ", "),
  Size_in_mem     = format(utils::object.size(df), "Mb"),
  N_countries     = length(unique(df$gwcode)),
  Years           = paste0(range(df$year, na.rm = TRUE), collapse = " - "),
  N_columns       = ncol(df),
  Columns         = paste0(colnames(df), collapse = ", "),
  N_rows          = nrow(df),
  N_complete_rows = sum(stats::complete.cases(df))
)
yaml::write_yaml(stats, "output/gdp-signature.yml")
stats

write_csv(joint, file = "output/gdp.csv")
```

