---
title: "Combined GDP data"
output: 
  github_document
---

*Last updated on: `r Sys.Date()`*

GDP data from 1950 on, mostly based on WDI, with some gaps filled with KSG's extended GDP data and data from the UN.  

Four step imputation procedure:

1. Acquire the WDI data
2. Where WDI is missing, drop in UN GDP figures, scaled by a linear model.
3. Where WDI is missing, drop in KSG figures, scaled by a log-linear country-varying scaling model.
4. Model-based extrapolation: use Kalman-smoothing to forward extrapolate missing GDP values (most notably Taiwan and several countries missing current year GDP values) and backward extrapolate GDP growth in first year of existences of a country.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

```{r}
library(dplyr)
library(ggplot2)
library(yaml)
library(tidyr)

gdp <- read.csv("output/gdp.csv")
str(gdp)
head(gdp)

stats <- yaml::read_yaml("output/gdp-signature.yml")
stats

gdp %>%
  pivot_longer(-one_of("gwcode", "year")) %>%
  ggplot(., aes(x = year, y = value, group = gwcode)) +
  facet_wrap(~ name, ncol = 1, scales = "free_y") +
  geom_line(alpha = .2) +
  theme_minimal()

```

